---
title: "Biological_Indicators"
output: html_document
date: "2025-09-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose of script: Calculate indicators for the NY Bight for DEC.

Contributors: Brandon Beltz, Laura Gruenburg
Last updated: 2025-09-10
Email: brandon.beltz@stonybrook.edu, laura.gruenburg@stonybrook.edu

```{r}
# set your working directory and load required libraries

setwd("/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/CalculateIndicators/Biological_indicators")

library(data.table); 
#library(rgdal); rgdal has been retired I am trying sf instead -LG 9/10/2025
library(survdat); 
library(tidyverse); 
library(ggplot2); 
library(dplyr); 
library(readr)
library(sf)

```

## Loading the Data

When I reached out to the NEFSC for the updated data in September of 2025 they gave me a file with everything in it so I have used that here.  Additionally we load the species code table and the strata shapefile so that we can pick the specific strata for the NYB

```{r}
## Load Survdat, species list and strata
data <- readRDS('surveyData_2025-09-10.rds')
#load('NEFSC_survey_2020_2022.RData')
#load('NEFSC_survey_2022_2023.RData')
load("Species_codes.RData")
strata<-read_sf('strata','strata')

#survey2 <- survey$survdat # extract the data from the most updated file
#survey3 <- survdat$survdat #ex
#load("Survdat.RData")
#dt = rbind(survdat[survdat$YEAR < 2020, ], survey2[survey2$YEAR <2022], survey3) # get rid of any over lapping data and merge old and new 

dt = data$survdat

```

## Calculate mean stratified biomass for FALL and SPRING

```{r}

## Generate area table
#strat.area<-get_area(strata, 'STRATA') # this did not work for Laura
#setnames(strat.area,'STRATA','STRATUM')

area <- sf::st_read(dsn=system.file("extdata","strata.shp",package="survdat"))
strat.area <- get_area(areaPolygon = area, areaDescription="STRATA")

## Identify strata
NYB.strata<-c(1010:1080)

## Subset by season and strata
NYB.fall<-dt[SEASON == 'FALL' & STRATUM %in% NYB.strata,]
NYB.spring<-dt[SEASON == 'SPRING' & STRATUM %in% NYB.strata,]

```
## Some of the survdat functions are not working from the library, I can't figure that out.  So I have reproduced them here just copied and pasted from Andy Beets lovely github page https://github.com/NOAA-EDAB/survdat/


## strat_prep

```{r}
#'Prepare survey data for calculating stratified mean
#'
#'Calculates the number of tows per strata and strata weight.  Run before
#'sub setting the data by species of interest to ensure all stations are accounted.
#'
#'
#' @param areaPolygon sf object or character string. Default = "NEFSC strata". The default option uses the survey strata shapefile bundled with the package.
#' To use any other shapefile for stratification, the shapefile must be read in as an \link[sf]{sf} object and the \code{areaDescription} argument must be specified.
#' @param areaDescription Character String. Column name from \code{areaPolygon}
#'                       that contains the strata designations.
#' @param surveyData Data table. NEFSC survey data generated by \code{get_survdat_data.R}
#' @param filterByArea Numeric vector. Set of areas to subset from the
#'  \code{areaDescription} of the \code{areaPolygon}.
#' @param filterBySeason Character string. Which seasons of the \code{surveyData}
#'  should be included.  Choices include "SPRING", "FALL", or "all".
#'
#' @return Returns a prepSurvey data object with added columns for the number of
#'  tows (ntows) and stratum weights (W.h).
#'
#' @importFrom data.table "key"
#'
#'@family survdat
#'
#'\@examples
#' \dontrun{
#' # Called internally
#' }
#'
#' @noRd


strat_prep <- function (surveyData, areaPolygon = "NEFSC strata",
                        areaDescription = "STRATA", filterByArea = "all",
                        filterBySeason = "all") {

  # Break link to original data set so no changes are made to original
  surveyData <- data.table::copy(surveyData)

  # Calculate the proportional areas
  # Use original stratified design and built-in shapefile or flag for post
  # stratification
  if(!is(areaPolygon, 'sf')){
    if(areaPolygon == 'NEFSC strata'){
      areaPolygon <- sf::st_read(dsn = system.file("extdata", "strata.shp",
                                                   package = "survdat"), quiet = T)
      poststratFlag <- F
    }
  } else poststratFlag <- T

  # Calculate area of the polygons
  polygonArea <- survdat::get_area(areaPolygon, areaDescription)

  # post stratify if necessary
  if(poststratFlag){
    message("Post stratifying ...")
    surveyData <- survdat:::post_strat(surveyData, areaPolygon, areaDescription)
  } else {
    #Add extra column to original data to mimic what happens when post-stratifying
    surveyData <- surveyData[, areaDescription := STRATUM]
    data.table::setnames(surveyData, 'areaDescription', areaDescription)
  }

  #FilterData
  if(filterByArea[1] != "all" | filterBySeason[1] != "all"){
    message("Filtering data ...")
    seasonFlag <- T
  }

  if(filterBySeason[1] == 'all'){
    filterBySeason <- unique(surveyData[, SEASON])
    seasonFlag <- F
  }

  # check to create all areas
  if (length(filterByArea)==1) {
    if (filterByArea == "all") {
      filterByArea <- areaPolygon %>%
        sf::st_drop_geometry() %>%
        dplyr::select(areaDescription) %>%
        unlist() %>%
        unique()
      if (is.factor(filterByArea)){ # convert to levels
        filterByArea <- levels(filterByArea)
      }
    }
  }
  print(filterByArea)

  filteredData <- surveyData[SEASON %in% filterBySeason & get(areaDescription) %in% filterByArea, ]

  #Change to generic names for calculations
  data.table::setnames(filteredData, areaDescription, 'STRAT')
  data.table::setnames(polygonArea, c("STRATUM", "Area"), c('STRAT', 'S.AREA'))

  # if not using original stratification need to preserve unique key
  if(!is.null(areaPolygon)){
    filteredData[, STATION2 := as.numeric(paste0(STRATUM, STATION))]
    data.table::setnames(filteredData, c('STATION', 'STATION2'),
                         c('ORIGSTATION', 'STATION'))
  }

  #Station data - Finds list of distinct stations sampled through time
  data.table::setkey(filteredData, CRUISE6, STRAT, STATION)
  stations <- unique(filteredData, by = key(filteredData))
  stations <- stations[, list(YEAR, CRUISE6, STRAT, STATION)]

  #x %>% dplyr::distinct(YEAR,CRUISE6,STRAT,STATION)
  # Count the number of stations in each year for each Region
  if(seasonFlag){
    data.table::setkey(stations, YEAR, CRUISE6, STRAT)
  } else { data.table::setkey(stations, YEAR, STRAT)}
  stations[, ntows := length(STATION), by = key(stations)]

  #Merge stations and area
  stations <- base::merge(stations, polygonArea, by = 'STRAT', all.x = T)

  #Calculate stratum weight
  if(seasonFlag){
    keyoff <- c('YEAR', 'CRUISE6')
  } else {
    keyoff <- 'YEAR'
  }
  data.table::setkeyv(stations, c(keyoff, 'STRAT'))
  strat.year <- unique(stations, by = key(stations))
  strat.year[, c('STATION', 'ntows') := NULL]
  strat.year[, W.h := S.AREA / sum(S.AREA, na.rm = T), by = keyoff]
  strat.year[, W.h := as.vector(W.h)] #Drops the units from the area
  strat.year[is.na(W.h), W.h := 0]
  strat.year[, S.AREA := NULL]
  if(!seasonFlag) strat.year[, CRUISE6 := NULL]

  #Merge back
  stations <- merge(stations, strat.year, by = key(stations))

  #Merge catch with station data
  prepData <- merge(filteredData, stations, by = c('YEAR', 'CRUISE6', 'STRAT', 'STATION'))

  data.table::setnames(prepData, c('STRAT', 'S.AREA'),
           c(areaDescription, "Area"))

  # Restore original station number if not using the original stratified design
  if(!is.null(areaPolygon)){
    data.table::setnames(prepData, c('STATION', 'ORIGSTATION'), c('STATION2', 'STATION'))
    prepData[, STATION2 := NULL]
  }

  return(prepData[])
}



```

## strat_mean

```{r}
#' Calculate stratified mean biomass and abundance
#'
#' Calculates the stratified mean biomass and abundance.  Also calculates the
#' associated variance and standard error.
#'
#'
#' @inheritParams strat_prep
#' @param prepData Data table. NEFSC survey data generated by \code{get_survdat_data}
#'  and modified by \code{strat_prep}.
#' @param groupDescription Character string. Column of \code{prepData} which
#'  contains the groups (e.g. "SVSPP") on which the means are based.
#' @param filterByGroup Character or numeric vector. Set of groups to subset from
#'  \code{groupDescription}.  The default "all" will calculate means for all groups.
#' @param mergesexFlag Boolean. Logical value to merge sexed species such as dogfish.
#' @param seasonFlag Boolean. Logical value to merge seasons (F) or keep them separate (T).
#' @param poststratFlag Boolean. Logical value indicating whether the original
#'  strata design was used or not.  Changes the calculation for the variance for
#'  post-stratified uses of survey data.
#'
#' @return Returns a table with stratified mean biomass and abundance for each group
#'  indicated by the \code{groupDescription}.  In addition, the variance and standard
#'  error for both means are provided.
#'
#' @importFrom data.table "key"
#'
#'@family surdat
#'
#'@examples
#'\dontrun{
#' # Called internally
#'}
#'
#' @noRd


strat_mean <- function (prepData, groupDescription = "SVSPP", filterByGroup = "all",
                        mergesexFlag = T, seasonFlag, areaDescription, poststratFlag) {

  stratmeanData <- data.table::copy(prepData)

  #Remove length data if present
  if(length(which(names(stratmeanData) == "LENGTH")) == 1){
    data.table::setkey(stratmeanData, CRUISE6, STRATUM, STATION, SVSPP, CATCHSEX)
    stratmeanData <- unique(stratmeanData, by = key(stratmeanData))
    stratmeanData[, c('LENGTH', 'NUMLEN') := NULL]
  }

  data.table::setnames(stratmeanData, c(groupDescription, areaDescription),
                       c('group', 'strat'))

  #Merge sex or keep separate
  if(mergesexFlag == F) stratmeanData[, group := paste(group, CATCHSEX, sep = '')]

  data.table::setkey(stratmeanData, CRUISE6, strat, STATION, group)
  stratmeanData[, BIOMASS   := sum(BIOMASS),   by = key(stratmeanData)]
  stratmeanData[, ABUNDANCE := sum(ABUNDANCE), by = key(stratmeanData)]
  stratmeanData <- unique(stratmeanData, by = key(stratmeanData))

  #Fix Na's
  stratmeanData[is.na(BIOMASS),   BIOMASS   := 0]
  stratmeanData[is.na(ABUNDANCE), ABUNDANCE := 0]

  #Calculate total number of stations per year
  #Determine if merging year or treating seasons separate
  if(seasonFlag){
    keyoff <- c('YEAR', 'SEASON')
  } else {
      keyoff <- 'YEAR'
    }
  data.table::setkey(stratmeanData, CRUISE6, strat, STATION)
  stations <- unique(stratmeanData, by = key(stratmeanData))
  N <- stations[, length(ntows), by = keyoff]
  data.table::setnames(N, 'V1', 'N')

  #Subset data if necessary
  if(filterByGroup[1] != 'all'){
    if(mergesexFlag == F) filterByGroup <- c(paste0(filterByGroup, 0),
                                             paste0(filterByGroup, 1),
                                             paste0(filterByGroup, 2),
                                             paste0(filterByGroup, 3))
    stratmeanData <- stratmeanData[group %in% filterByGroup, ]
  }

  #Calculate weight per tow and number per tow
  data.table::setkeyv(stratmeanData, c('group', 'strat', keyoff))

  stratmeanData[, biomass.tow   := sum(BIOMASS)   / ntows, by = key(stratmeanData)]
  stratmeanData[, abundance.tow := sum(ABUNDANCE) / ntows, by = key(stratmeanData)]

  #Calculated stratified means
  stratmeanData[, weighted.biomass   := biomass.tow   * W.h]
  stratmeanData[, weighted.abundance := abundance.tow * W.h]

  #Variance - need to account for zero catch
  stratmeanData[, n.zero := ntows - length(BIOMASS), by = key(stratmeanData)]

  stratmeanData[, zero.var.b := n.zero * (0 - biomass.tow)^2]
  stratmeanData[, vari.b := (BIOMASS - biomass.tow)^2]
  stratmeanData[, Sh.2.b := (zero.var.b + sum(vari.b)) / (ntows - 1), by = key(stratmeanData)]
  stratmeanData[is.nan(Sh.2.b), Sh.2.b := 0]

  stratmeanData[, zero.var.a := n.zero * (0 - abundance.tow)^2]
  stratmeanData[, vari.a := (ABUNDANCE - abundance.tow)^2]
  stratmeanData[, Sh.2.a := (zero.var.a + sum(vari.a)) / (ntows - 1), by = key(stratmeanData)]
  stratmeanData[is.nan(Sh.2.a), Sh.2.a := 0]

  stratmeanData <- unique(stratmeanData, by = key(stratmeanData))

  stratmeanData <- merge(stratmeanData, N, by = keyoff)

  #Stratified mean
  data.table::setkeyv(stratmeanData, c('group', keyoff))

  stratmeanData[, strat.biomass := sum(weighted.biomass),   by = key(stratmeanData)]
  stratmeanData[, strat.abund   := sum(weighted.abundance), by = key(stratmeanData)]

  #Stratified variance
  if(poststratFlag == F){
    stratmeanData[, biomass.var := sum(((W.h^2) * Sh.2.b) / ntows), by = key(stratmeanData)]
    stratmeanData[, abund.var   := sum(((W.h^2) * Sh.2.a) / ntows), by = key(stratmeanData)]
  }

  if(poststratFlag == T){
    stratmeanData[, biomass.var := sum(Sh.2.b * W.h) / N + sum((1 - W.h) * Sh.2.b) / N^2, by = key(stratmeanData)]
    stratmeanData[, abund.var   := sum(Sh.2.a * W.h) / N + sum((1 - W.h) * Sh.2.a) / N^2, by = key(stratmeanData)]

  }

  #standard error of the means
  stratmeanData[, biomass.SE := sqrt(biomass.var), by = key(stratmeanData)]
  stratmeanData[, abund.SE   := sqrt(abund.var),   by = key(stratmeanData)]

  #Delete extra rows/columns
  stratmeanData <- unique(stratmeanData, by = key(stratmeanData))
  stratmeanData <- stratmeanData[, list(YEAR, SEASON, group, CATCHSEX, N, strat.biomass,
                                        biomass.var, biomass.SE, strat.abund,
                                        abund.var, abund.SE)]
  if(mergesexFlag == T) stratmeanData[, CATCHSEX := NULL]

  if(mergesexFlag == F){
    stratmeanData[, glen := nchar(group)]
    for(i in 2:4){
      stratmeanData[glen == i, group := as.numeric(substr(group, 1, i - 1))]
    }
    stratmeanData[, glen := NULL]
    data.table::setkey(stratmeanData, YEAR, SVSPP, CATCHSEX)
  }

  if(seasonFlag == F) stratmeanData[, SEASON := 'ALL']

  data.table::setnames(stratmeanData, 'group', groupDescription)

  return(stratmeanData[])
}

```

## swept_area

```{r}
#'Calculate the estimate of total biomass and abundance
#'
#'Calculates the estimate of total biomass and abundance using the swept area.
#'Variance of the estimate is also calculated.  Requires the outputs from both
#'the stratprep and stratmean functions.
#'
#'
#'@param survdat NEFSC survey data generated by Survdat.R and modified by stratprep().
#'@param stratmean Output from strat_mean().
#'@param q Table of survey catchability with a column of group names and a column of
#'catchabilities.  If not provided, assumes a q of 1 for each group (Minimum swept area
#'estimates).
#'@param a The average swept area of the trawl.  Default value is the swept area of a
#'standard NOAA Ship Albatross IV tow.
#'@param areaDescription Column of survdat containing strata designations.
#'@param groupDescription Column of survdat upon which the totals are based (i.e. SVSPP).
#'
#'@return Returns a table with the estimates of total biomass and abundance as well as
#'the stratified mean biomass and abundance for each group indicated by the groupDescription
#'parameter.  In addition, the variance of the totals and variance and standard error
#'for both means are provided.
#'
#'@importFrom data.table "key"
#'
#'@family survdat
#'
#'@examples
#'\dontrun{
#'# Called internally
#'}
#'
#'@noRd

swept_area <- function(
  prepData,
  stratmeanData,
  q = NULL,
  a = 0.0384,
  areaDescription,
  groupDescription = 'SVSPP'
) {
  #This is necessary to break the link with the original data table
  prepData.x <- data.table::copy(prepData)
  stratmeanData.x <- data.table::copy(stratmeanData)

  #Calculate A (Total area)
  data.table::setnames(
    prepData.x,
    c(areaDescription, "Area"),
    c('STRAT', 'S.AREA')
  )

  data.table::setkey(prepData.x, YEAR, STRAT)
  stratum <- unique(prepData.x, by = key(prepData.x))
  stratum <- stratum[, sum(S.AREA, na.rm = T), by = 'YEAR']
  data.table::setnames(stratum, "V1", "A")

  #Merge A
  swept.area <- base::merge(stratmeanData.x, stratum, by = 'YEAR')

  #Merge q
  if (is.null(q)) {
    q <- data.table::data.table(
      groups = unique(swept.area[, get(groupDescription)]),
      q = 1
    )
    message("Assuming a value of q = ", q, " for all groups")
  } else if (length(q) == 1) {
    # apply q to all groups
    message("Assuming a value of q = ", q, " for all groups")
    q <- data.table::data.table(
      groups = unique(swept.area[, get(groupDescription)]),
      q = q
    )
  } else {
    # Need a value of q for all groups
  }

  data.table::setnames(q, names(q), c(groupDescription, 'q'))
  swept.area <- base::merge(swept.area, q, by = groupDescription, all.x = T)
  swept.area[is.na(q), q := 1]

  #Calculate swept area biomass
  swept.area[, tot.biomass := (strat.biomass * A / a) / q]
  swept.area[, tot.abundance := round((strat.abund * A / a) / q)]

  #Calculate variance
  swept.area[, var.constant := (A / a) / q]
  swept.area[, tot.bio.var := var.constant^2 * biomass.var]
  swept.area[, tot.abund.var := var.constant^2 * abund.var]

  #Calculate standard error
  swept.area[, tot.bio.SE := sqrt(tot.bio.var)]
  swept.area[, tot.abund.SE := sqrt(tot.abund.var)]

  #remove extra columns - need to add sex column if stratmean object does not have one
  #then remove before output
  if (length(which(names(stratmeanData.x) == 'sex')) == 0) {
    swept.area[, sex := 0]
  }
  swept.area <- swept.area[, list(
    YEAR,
    get(groupDescription),
    sex,
    N,
    strat.biomass,
    biomass.var,
    biomass.SE,
    strat.abund,
    abund.var,
    abund.SE,
    tot.biomass,
    tot.bio.var,
    tot.bio.SE,
    tot.abundance,
    tot.abund.var,
    tot.abund.SE
  )]
  data.table::setnames(swept.area, 'V2', groupDescription)
  if (length(which(names(stratmeanData.x) == 'sex')) == 0) {
    swept.area[, sex := NULL]
  }

  swept.area <- swept.area %>% units::drop_units()

  return(swept.area)
}

```

```{r}
## Run stratification prep
NYB.prep.fall<-strat_prep(NYB.fall,areaPolygon = area, areaDescription="STRATA")#,strat.col = 'STRATUM',area.col = 'Area')
NYB.prep.spring<-strat_prep(NYB.spring,areaPolygon = area, areaDescription="STRATA")

## Calculate stratified means
mean.biomass.fall<-strat_mean(NYB.prep.fall,groupDescription = 'SVSPP', areaDescription = 'STRATA',seasonFlag = T,poststratFlag = T)#group.col = 'SVSPP',strat.col = 'STRATUM')
mean.biomass.spring<-strat_mean(NYB.prep.spring,groupDescription = 'SVSPP', areaDescription = 'STRATA',seasonFlag = T,poststratFlag = T)

## Merge with RPATH names
mean.biomass.fall<-merge(mean.biomass.fall,spp[,list(SVSPP,RPATH,SCINAME,Fall.q)], by = 'SVSPP')
mean.biomass.spring<-merge(mean.biomass.spring,spp[,list(SVSPP,RPATH,SCINAME,Spring.q)], by = 'SVSPP')

## Calculate total biomass from swept area
total.biomass.fall<-swept_area(NYB.prep.fall, mean.biomass.fall, groupDescription = 'SVSPP', areaDescription = 'STRATA')#, strat.col = 'STRATUM', area.col = 'Area')
total.biomass.spring<-swept_area(NYB.prep.spring, mean.biomass.spring, groupDescription = 'SVSPP', areaDescription = 'STRATA')

## Merge with RPATH names
total.biomass.fall<-merge(total.biomass.fall,spp[,list(SVSPP,RPATH,SCINAME,Fall.q)], by = 'SVSPP')
total.biomass.spring<-merge(total.biomass.spring,spp[,list(SVSPP,RPATH,SCINAME,Spring.q)], by = 'SVSPP')

## Calculate total area
NYB.strat.area<-strat.area[STRATUM %in% NYB.strata,sum(Area)]

## Convert to b/a in mt
total.biomass.fall<-total.biomass.fall[, biomass.t_area :=(tot.biomass*.001)/(Fall.q*NYB.strat.area)]
total.biomass.spring<-total.biomass.spring[, biomass.t_area :=(tot.biomass*.001)/(Spring.q*NYB.strat.area)]

##Output to .csv and .RData
save(total.biomass.fall, file = 'NYB_biomass_fall_2025.RData')
write.csv(total.biomass.fall, file = "NYB_biomass_fall2025.csv")

save(total.biomass.spring, file = 'NYB_biomass_spring_2025.RData')
write.csv(total.biomass.spring, file = "NYB_biomass_spring_2025.csv")
```

## Creating and saving time series for each indicator

```{r}
# Longfin squid biomass
longfin.fall<-subset(total.biomass.fall, SCINAME == 'LOLIGO PEALEII',select = c("YEAR","strat.biomass"))
longfin.spring<-subset(total.biomass.spring, SCINAME == 'LOLIGO PEALEII',select = c("YEAR","strat.biomass"))
write.csv(longfin.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/longfin_fall_2025.csv")
write.csv(longfin.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/longfin_spring_2025.csv")

# Northern shortfin squid biomass
shortfin.fall<-subset(total.biomass.fall, SCINAME == 'ILLEX ILLECEBROSUS',select = c("YEAR","strat.biomass"))
shortfin.spring<-subset(total.biomass.spring, SCINAME == 'ILLEX ILLECEBROSUS',select = c("YEAR","strat.biomass"))
write.csv(shortfin.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/shortfin_fall_2025.csv")
write.csv(shortfin.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/shortfin_spring_2025.csv")

# Black sea bass biomass
bsb.fall<-subset(total.biomass.fall, SCINAME == 'CENTROPRISTIS STRIATA',select = c("YEAR","strat.biomass"))
bsb.spring<-subset(total.biomass.spring, SCINAME == 'CENTROPRISTIS STRIATA',select = c("YEAR","strat.biomass"))
write.csv(bsb.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/bsb_fall_2025.csv")
write.csv(bsb.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/bsb_spring_2025.csv")

# Summer flounder biomass
summerflounder.fall<-subset(total.biomass.fall, SCINAME == 'PARALICHTHYS DENTATUS',select = c("YEAR","strat.biomass"))
summerflounder.spring<-subset(total.biomass.spring, SCINAME == 'PARALICHTHYS DENTATUS',select = c("YEAR","strat.biomass"))
write.csv(summerflounder.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/summerflounder_fall_2025.csv")
write.csv(summerflounder.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/summerflounder_spring_2025.csv")

# Jonah crab biomass
jonah.fall<-subset(total.biomass.fall, SCINAME == 'CANCER BOREALIS',select = c("YEAR","strat.biomass"))
jonah.spring<-subset(total.biomass.spring, SCINAME == 'CANCER BOREALIS',select = c("YEAR","strat.biomass"))
write.csv(jonah.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/jonah_fall_2025.csv")
write.csv(jonah.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/jonah_spring_2025.csv")

# American lobster biomass
amlob.fall<-subset(total.biomass.fall, SCINAME == 'HOMARUS AMERICANUS',select = c("YEAR","strat.biomass"))
amlob.spring<-subset(total.biomass.spring, SCINAME == 'HOMARUS AMERICANUS',select = c("YEAR","strat.biomass"))
write.csv(amlob.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/amlob_fall_2025.csv")
write.csv(amlob.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/amlob_spring_2025.csv")

# Total trawl biomass
trawl.fall<-total.biomass.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(trawl.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/trawl_fall_2025.csv")

trawl.spring<-total.biomass.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(trawl.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/trawl_spring_2025.csv")

# Assign species preferences from Janet
species_prefs <- read_csv("species_prefs.csv")
total.biomass.fall<-merge(total.biomass.fall,species_prefs,by='SVSPP')
total.biomass.spring<-merge(total.biomass.spring,species_prefs,by='SVSPP')

# Northern v Summer species
north.fall<-subset(total.biomass.fall, NORTH.SOUTH == 'N',select = c("YEAR","strat.biomass"))
north.fall<-north.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(north.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/north_fall_2025.csv")

north.spring<-subset(total.biomass.spring, NORTH.SOUTH == 'N',select = c("YEAR","strat.biomass"))
north.spring<-north.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(north.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/north_spring_2025.csv")

south.fall<-subset(total.biomass.fall, NORTH.SOUTH == 'S',select = c("YEAR","strat.biomass"))
south.fall<-south.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(south.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/south_fall_2025.csv")

south.spring<-subset(total.biomass.spring, NORTH.SOUTH == 'S',select = c("YEAR","strat.biomass"))
south.spring<-south.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(south.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/south_spring_2025.csv")

# Benthic v Pelagic species
benthic.fall<-subset(total.biomass.fall, BENTH.PEL == 'Benthic',select = c("YEAR","strat.biomass"))
benthic.fall<-benthic.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthic.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthic_fall_2025.csv")

benthic.spring<-subset(total.biomass.spring, BENTH.PEL == 'Benthic',select = c("YEAR","strat.biomass"))
benthic.spring<-benthic.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthic.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthic_spring_2025.csv")

pelagic.fall<-subset(total.biomass.fall, BENTH.PEL == 'Pelagic',select = c("YEAR","strat.biomass"))
pelagic.fall<-pelagic.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(pelagic.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/pelagic_fall_2025.csv")

pelagic.spring<-subset(total.biomass.spring, BENTH.PEL == 'Pelagic',select = c("YEAR","strat.biomass"))
pelagic.spring<-pelagic.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(pelagic.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/pelagic_spring_2025.csv")

# Average TL of community by year
avgtl.fall<-subset(total.biomass.fall, SVSPP != '950')
avgtl.fall<-avgtl.fall %>%
  group_by(YEAR) %>%
  summarise(TL = weighted.mean(TL,strat.biomass))
write.csv(avgtl.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/avgtl_fall_2025.csv")

avgtl.spring<-subset(total.biomass.spring, SVSPP != '950')
avgtl.spring<-avgtl.spring %>%
  group_by(YEAR) %>%
  summarise(TL = weighted.mean(TL,strat.biomass))
write.csv(avgtl.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/avgtl_spring_2025.csv")

# Average T preference of community by year
avgtemp.fall<-subset(total.biomass.fall, wt_mean_temp != 'NA')
avgtemp.fall<-avgtemp.fall %>%
  group_by(YEAR) %>%
  summarise(wt_mean_temp = weighted.mean(wt_mean_temp,strat.biomass))
write.csv(avgtemp.fall,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/avgtemp_fall_2025.csv")

avgtemp.spring<-subset(total.biomass.spring, wt_mean_temp != 'NA')
avgtemp.spring<-avgtemp.spring %>%
  group_by(YEAR) %>%
  summarise(wt_mean_temp = weighted.mean(wt_mean_temp,strat.biomass))
write.csv(avgtemp.spring,file = "/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/avgtemp_spring_2025.csv")

# Assign NEIEA categories
neiea_spp <- read_csv("forage_spp.csv")
total.biomass.fall<-merge(total.biomass.fall,neiea_spp,by='SVSPP')
total.biomass.spring<-merge(total.biomass.spring,neiea_spp,by='SVSPP')

# Forage fish species biomass
forage.fall<-subset(total.biomass.fall, NEIEA == 'Forage Fish',select = c("YEAR","strat.biomass"))
forage.fall<-forage.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(forage.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/forage_fall_2025.csv")

forage.spring<-subset(total.biomass.spring, NEIEA == 'Forage Fish',select = c("YEAR","strat.biomass"))
forage.spring<-forage.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(forage.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/forage_spring_2025.csv")

# Assign NOAA feeding guilds
noaa_guilds <- read_csv("noaa_guilds.csv")
total.biomass.fall<-merge(total.biomass.fall,noaa_guilds,by='SVSPP')
total.biomass.spring<-merge(total.biomass.spring,noaa_guilds,by='SVSPP')

# Benthos biomass
benthos.fall<-subset(total.biomass.fall, Feeding.guild == 'Benthos',select = c("YEAR","strat.biomass"))
benthos.fall<-benthos.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthos.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthos_fall_2025.csv")

benthos.spring<-subset(total.biomass.spring, Feeding.guild == 'Benthos',select = c("YEAR","strat.biomass"))
benthos.spring<-benthos.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthos.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthos_spring_2025.csv")

# Benthivore biomass
benthivore.fall<-subset(total.biomass.fall, Feeding.guild == 'Benthivore',select = c("YEAR","strat.biomass"))
benthivore.fall<-benthivore.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthivore.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthivore_fall_2025.csv")

benthivore.spring<-subset(total.biomass.spring, Feeding.guild == 'Benthivore',select = c("YEAR","strat.biomass"))
benthivore.spring<-benthivore.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(benthivore.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/benthivore_spring_2025.csv")

# Planktivore biomass
planktivore.fall<-subset(total.biomass.fall, Feeding.guild == 'Planktivore',select = c("YEAR","strat.biomass"))
planktivore.fall<-planktivore.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(planktivore.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/planktivore_fall_2025.csv")

planktivore.spring<-subset(total.biomass.spring, Feeding.guild == 'Planktivore',select = c("YEAR","strat.biomass"))
planktivore.spring<-planktivore.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(planktivore.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/planktivore_spring_2025.csv")

# Piscivore biomass
piscivore.fall<-subset(total.biomass.fall, Feeding.guild == 'Piscivore',select = c("YEAR","strat.biomass"))
piscivore.fall<-piscivore.fall %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(piscivore.fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/piscivore_fall_2025.csv")

piscivore.spring<-subset(total.biomass.spring, Feeding.guild == 'Piscivore',select = c("YEAR","strat.biomass"))
piscivore.spring<-piscivore.spring %>%
  group_by(YEAR) %>%
  summarise(strat.biomass = sum(strat.biomass))
write.csv(piscivore.spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/piscivore_spring_2025.csv")

# Indicate which species are fish
fish_spp<-read_csv("fish_spp.csv")
total.biomass.fall<-merge(total.biomass.fall,fish_spp,by='SVSPP')
total.biomass.spring<-merge(total.biomass.spring,fish_spp,by='SVSPP')

# Fish species richness
library(dplyr)
unique_spp_fall<-subset(total.biomass.fall, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_fall<-distinct(unique_spp_fall)
unique_spp_fall<-count(unique_spp_fall, YEAR)
write.csv(unique_spp_fall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/richness_fall_2025.csv")

unique_spp_spring<-subset(total.biomass.spring, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_spring<-distinct(unique_spp_spring)
unique_spp_spring<-count(unique_spp_spring, YEAR)
write.csv(unique_spp_spring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/richness_spring_2025.csv")

# Richness Northern and Southern
north.fall <- total.biomass.fall[total.biomass.fall$NORTH.SOUTH == 'N', ]
south.fall <- total.biomass.fall[total.biomass.fall$NORTH.SOUTH == 'S', ]

north.spring <- total.biomass.spring[total.biomass.spring$NORTH.SOUTH == 'N', ]
south.spring <- total.biomass.spring[total.biomass.spring$NORTH.SOUTH == 'S']

unique_spp_Nfall<-subset(north.fall, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_Nfall<-distinct(unique_spp_Nfall)
unique_spp_Nfall<-count(unique_spp_Nfall, YEAR)
write.csv(unique_spp_Nfall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/Nrichness_fall_2025.csv")

unique_spp_Sfall<-subset(south.fall, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_Sfall<-distinct(unique_spp_Sfall)
unique_spp_Sfall<-count(unique_spp_Sfall, YEAR)
write.csv(unique_spp_Sfall,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/Srichness_fall_2025.csv")

unique_spp_Nspring<-subset(north.spring, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_Nspring<-distinct(unique_spp_Nspring)
unique_spp_Nspring<-count(unique_spp_Nspring, YEAR)
write.csv(unique_spp_Nspring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/Nrichness_spring_2025.csv")

unique_spp_Sspring<-subset(south.spring, Fish == 'Y',select = c("SVSPP","YEAR"))
unique_spp_Sspring<-distinct(unique_spp_Sspring)
unique_spp_Sspring<-count(unique_spp_Sspring, YEAR)
write.csv(unique_spp_Sspring,file="/Users/lauragruenburg/Desktop/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2025/Srichness_spring_2025.csv")


detach(package:dplyr)
```
