---
title: "Bottom_DO"
author: "laura gruenburg"
date: '2024-01-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the data

```{r}
library(dplyr)
library(tidyr)

#Ask Boashan for any updated bottle data first!
setwd('/Users/nyelab/Desktop/NYB Indicators/NYB_Indicators_Calculations')
do <- read.csv('NYOS_niskin_bottle_data_all_20231221.csv')
bot_do$Station <- as.numeric(bot_do$Station)

bot_do <- unique(bot_do[!is.na(bot_do$Station),]$Station)

bot_do <- do[do$Niskin_Bottle == 1, ]

bottomDO <- bot_do %>% group_by(Cruise) %>%
  summarise(Year = mean(Year),
            Month = mean(Month),
            DO = mean(DO..mg.L.))

season <- c('W', 'Sp','Su', 'F',
            'W', 'Sp', 'F',
            'W','Sp',
            'W','Sp','Su','F',
            'Sp', 'F'
            )

bottomDO$Season <- season

#Data from 202006 seems wrong.  We excluded it previously so I will do the same here
bottomDO <- bottomDO[!bottomDO$Cruise == 'NYOS2006',]

setwd('/Users/nyelab/Desktop/NYB Indicators/NYB_Indicators_Calculations/Final_Timeseries_Figures/Timeseries_Files_2023/')
write.csv(bottomDO, 'Bottom_DO_2023.csv')

bottomDO