---
title: "Surface_salinity_2023"
author: "laura gruenburg"
date: '2023-12-29'
output: html_document
---

Load required packages
```{r}

library(rgdal)
library(lubridate)
library(mgcv)
library(effects)
#library(mgcViz)
library(stringr)
#####load required functiosn and shapefiles
setwd("~/Desktop/NYB Indicators/CalculateIndicators/Rfunctions")
source("LabelPoints.R")

```

Load and combine the wodb and seawolf data
```{r}
#######Load the datasets
#######Load the datasets
setwd("~/Desktop/NYB Indicators/CalculateIndicators/WaterTemperature/Data")
ddd<-read.csv("WOD_CTD_format_D50_June24_2020.csv", header = TRUE)
ddd$date<-as.Date(ddd$date)
ddd$year<-format(ddd$date, format="%Y")
ddd$data_source<-"WOD"

setwd("~/Desktop/NYB Indicators/CalculateIndicators/Salinity")
ddd2<-read.csv("WOD_CTD_Dec29salinity_2023.csv")
colnames(ddd2)[18] <- "bot_sal"
colnames(ddd2)[16] <- "surf_sal"
ddd2$date <- as.character(ddd2$date)
ddd2$date<-as.Date(ddd2$date,tryFormats = c("%Y%m%d"))
ddd2$date<-as.Date(ddd2$date)
ddd2$year<-format(ddd2$date, format="%Y")
ddd2$month<-month(ymd(ddd2$date))
ddd2$data_source<-"WOD"

setwd("~/Desktop/NYB Indicators/CalculateIndicators/WaterTemperature/Data")
sea<-read.csv("CTD_seawolf_DEC12_2022.csv")
sea$date<-as.Date(sea$date)
sea$data_source<-"SEAWOLF"

sea2<-read.csv("CTD_seawolf_DEC5_2023.csv")
sea2$date<-as.Date(sea2$date)
sea2$data_source<-"SEAWOLF"

ddd2018 <- ddd[ddd$year < 2019, ] #update the last 5 years
common_cols <- intersect(colnames(ddd2018), colnames(ddd2))
ddd_combined<-rbind(
  subset(ddd2018, select = common_cols), 
  subset(ddd2, select = common_cols)
)

sea_combined<- rbind(sea,sea2)

common_cols <- intersect(colnames(ddd_combined), colnames(sea_combined))
ds<-rbind(
  subset(ddd_combined, select = common_cols), 
  subset(sea_combined, select = common_cols)
)

```


Format the combined data
```{r}

#Format date
ds$date<-as.Date(ds$date)
ds$year<-year(ds$date)
ds$day<-yday(ds$date)
ds$month<-month(ds$date)
ds$day<-yday(ds$date)
ds$year_fac<-factor(ds$year)



###use the function to label
lazylab<-klab(ds$lat, ds$lon)
ndf<-data.frame(ds, lazylab)
ndf$EPU<-as.character(ndf$EPU)
###Label casts not in an EPU
ndf[is.na(ndf$EPU),"EPU"]<-"Not in NES"

ndf$surf_sal[ndf$surf_sal<0] <- NA #some values are -99.99 which doesn't make sense - probably a filler for bad/missing data
ndf$surf_sal[ndf$surf_sal>40] <- NA #some values are 99.99 or over 100 which doesn't make sense - probably a filler for bad/missing data

ndf$bot_sal[ndf$bot_sal<0] <- NA #some values are -99.99 which doesn't make sense - probably a filler for bad/missing data
ndf$bot_sal[ndf$bot_sal>40] <- NA #some values are 99.99 or over 100 which doesn't make sense - probably a filler for bad/missing data

###remove values that are funky
hist(ndf$surf_sal)#guess some are pretty fresh
hist(ndf$bot_sal)#guess some are pretty fresh

```


#Start Analysis for the NYB
```{r}
########ANALYSIS FOR THE NYB##########
NYB<-ndf[ndf$NYB %in% "NYB" == TRUE,]
nrow(NYB)##9251 casts
minN<-10##min number of obs per year (you choose) to retain data for an estimate. 
ntab<-table(NYB$year_fac)#make a table
ntab
remove_these<-names(ntab[ntab < minN])#remove below minN
###subdat1 removes data for years with less than minNcasts and is used in analysis
subdat1<-NYB[NYB$year_fac %in% remove_these == FALSE,]
nrow(subdat1)#this is how many casts are used (9241)




######Display Data Density + TRIM
table(subdat1$year)#casts per year for data used in analysis
table(NYB$year)#cast per year for whole NYB
#plot of each year with number of casts per month
par(mfrow=c(2,2))
yearz<-unique(NYB$year)
for(i in 1:length(yearz)){
  #i = 1
  daYear<-NYB[NYB$year == yearz[i],]
  barplot(table(daYear$month), main = paste(yearz[i], " ", nrow(daYear)))
}
dev.off()

#seawolf data vs. WOD data
table(NYB$data_source, NYB$year)
table(NYB$data_source)

```


Seasonal
```{r}

# seasonal bottom temp
subdat1$season<-NA
subdat1[subdat1$day %in% 1:90 == TRUE,"season"]<-"winter"
subdat1[subdat1$day %in% 91:181 == TRUE,"season"]<-"spring"
subdat1[subdat1$day %in% 182:273 == TRUE,"season"]<-"summer"
subdat1[subdat1$day %in% 274:366 == TRUE,"season"]<-"fall"
table(subdat1$season)

# WINTER
winter <- subdat1[subdat1$season == 'winter', ]
NYB_winter<-gam(surf_sal ~ s(lon,lat, k = 100)  +  year_fac,  data = winter)
summary(NYB_winter)#check out model
gam.check(NYB_winter)
par(mfrow = c(1,2))
plot(NYB_winter)#month effect is interesting

plot(NYB_winter$fitted.values, NYB_winter$y, pch = 19, cex = .3,
     col = factor(winter$data_source))
abline(0,1)

regexp <- "[[:digit:]]+"
index_vals<-NYB_winter$coefficients[grepl("year",names(NYB_winter$coefficients))]
intercept_winter<- NYB_winter$coefficients[grepl("(Intercept)",names(NYB_winter$coefficients))]
years<-as.numeric(str_extract(names(index_vals), regexp))
index_1<-data.frame(years, index_vals)#THe first year of the data is the baseline
#so need to add in first year as a 0
indexNYB_winter<-rbind(data.frame(years = min(winter$year, na.rm = TRUE),index_vals = 0), index_1)
fNYB_winter<-data.frame(Year = indexNYB_winter$years,
                 Variable = "SurfSal_insitu",
                 Intercept = intercept_winter,
                 Val = scale(indexNYB_winter$index_vals, scale = FALSE),
                 Loc = "NYB",
                 N = as.vector(table(winter$year)))
plot(fNYB_winter$Year, scale(fNYB_winter$Val, scale = FALSE), type = "b")
plot(fNYB_winter$Year,fNYB_winter$Intercept + fNYB_winter$Val, type = "b")
abline(h = 0)
abline(v=2003)

# spring
spring <- subdat1[subdat1$season == 'spring', ]
NYB_spring<-gam(surf_sal ~ s(lon,lat, k = 100)  +  year_fac,  data = spring)
summary(NYB_spring)#check out model
gam.check(NYB_spring)
par(mfrow = c(1,2))
plot(NYB_spring)#month effect is interesting

plot(NYB_spring$fitted.values, NYB_spring$y, pch = 19, cex = .3,
     col = factor(spring$data_source))
abline(0,1)

regexp <- "[[:digit:]]+"
index_vals<-NYB_spring$coefficients[grepl("year",names(NYB_spring$coefficients))]
intercept_spring<- NYB_spring$coefficients[grepl("(Intercept)",names(NYB_spring$coefficients))]
years<-as.numeric(str_extract(names(index_vals), regexp))
index_1<-data.frame(years, index_vals)#THe first year of the data is the baseline
#so need to add in first year as a 0
indexNYB_spring<-rbind(data.frame(years = min(spring$year, na.rm = TRUE),index_vals = 0), index_1)
fNYB_spring<-data.frame(Year = indexNYB_spring$years,
                        Variable = "SurfSal_insitu",
                        Intercept = intercept_spring,
                        Val = scale(indexNYB_spring$index_vals, scale = FALSE),
                        Loc = "NYB",
                        N = as.vector(table(spring$year)))
plot(fNYB_spring$Year, scale(fNYB_spring$Val, scale = FALSE), type = "b")
plot(fNYB_spring$Year, fNYB_spring$Intercept + fNYB_spring$Val, type = 'b')
abline(h = 0)
abline(v=2003)

# summer
summer <- subdat1[subdat1$season == 'summer', ]
NYB_summer<-gam(surf_sal ~ s(lon,lat, k = 100)  +  year_fac,  data = summer)
summary(NYB_summer)#check out model
gam.check(NYB_summer)
par(mfrow = c(1,2))
plot(NYB_summer)#month effect is interesting

plot(NYB_summer$fitted.values, NYB_summer$y, pch = 19, cex = .3,
     col = factor(summer$data_source))
abline(0,1)

regexp <- "[[:digit:]]+"
index_vals<-NYB_summer$coefficients[grepl("year",names(NYB_summer$coefficients))]
intercept_summer<- NYB_summer$coefficients[grepl("(Intercept)",names(NYB_summer$coefficients))]
years<-as.numeric(str_extract(names(index_vals), regexp))
index_1<-data.frame(years, index_vals)#THe first year of the data is the baseline
#so need to add in first year as a 0
indexNYB_summer<-rbind(data.frame(years = min(summer$year, na.rm = TRUE),index_vals = 0), index_1)
fNYB_summer<-data.frame(Year = indexNYB_summer$years,
                        Variable = "SurfSal_insitu",
                        Intercept = intercept_summer,
                        Val = scale(indexNYB_summer$index_vals, scale = FALSE),
                        Loc = "NYB",
                        N = as.vector(table(summer$year)))
plot(fNYB_summer$Year, scale(fNYB_summer$Val, scale = FALSE), type = "b")
plot(fNYB_summer$Year, fNYB_summer$Intercept + fNYB_summer$Val, type = "b")
abline(h = 0)
abline(v=2003)

# fall
fall <- subdat1[subdat1$season == 'fall', ]
NYB_fall<-gam(surf_sal ~ s(lon,lat, k = 100)  +  year_fac,  data = fall)
summary(NYB_fall)#check out model
gam.check(NYB_fall)
par(mfrow = c(1,2))
plot(NYB_fall)#month effect is interesting

plot(NYB_fall$fitted.values, NYB_fall$y, pch = 19, cex = .3,
     col = factor(fall$data_source))
abline(0,1)

regexp <- "[[:digit:]]+"
index_vals<-NYB_fall$coefficients[grepl("year",names(NYB_fall$coefficients))]
intercept_fall<- NYB_fall$coefficients[grepl("(Intercept)",names(NYB_fall$coefficients))]
years<-as.numeric(str_extract(names(index_vals), regexp))
index_1<-data.frame(years, index_vals)#THe first year of the data is the baseline
#so need to add in first year as a 0
indexNYB_fall<-rbind(data.frame(years = min(fall$year, na.rm = TRUE),index_vals = 0), index_1)
fNYB_fall<-data.frame(Year = indexNYB_fall$years,
                        Variable = "SurfSal_insitu",
                        Intercept = intercept_fall,
                        Val = scale(indexNYB_fall$index_vals, scale = FALSE),
                        Loc = "NYB",
                        N = as.vector(table(fall$year)))
plot(fNYB_fall$Year, scale(fNYB_fall$Val, scale = FALSE), type = "b")
plot(fNYB_fall$Year, fNYB_fall$Intercept + fNYB_fall$Val, type = 'b')
abline(h = 0)
abline(v=2003)

fNYB_winter$season<-'winter'
fNYB_spring$season<-'spring'
fNYB_summer$season<-'summer'
fNYB_fall$season<-'fall'

fNYB <- rbind(fNYB_winter, fNYB_spring, fNYB_summer, fNYB_fall)

setwd("~/Desktop/NYB Indicators/NYB_Indicators_Calculations/Final_timeseries_figures/Timeseries_Files_2023")
write.csv(fNYB, "SurfSal_insitu_Dec_29_2023.csv")




